---
title: "Regression Class Project"
author: "_Winston Saunders_ November 2014"
output: pdf_document
---

##Executive Summary

This analysis looks at factors affecting the efficiency (__mpg__) of cars in the Motor Trend data set. Particularly, it addresses the hypothesis that manual transmissions are more efficient than automatic transmissions. 

The conclusion of the analysis is that we cannot reject the null hypothesis. There is no significan difference, for the cars in this data set, in the miles per gallon of automatic versus manual transmissions.


##Exploratory data analysis


```{r echo=FALSE}
data(mtcars)
library(plyr)
library(ggplot2)

##clean up factor
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <-c("Automatic", "Manual")


##create new data columns for inverse weight adn displacement


mtcars2<-mutate(mtcars, invwt=1/wt)
mtcars2<-mutate(mtcars2, invdisp=1/disp)


```
  

There are data for `r nrow(mtcars)` cars in the __mtcars__ data set representing _`r colnames(mtcars)`_.  Of interest here are:
__mpg__ - Miles/(US) gallon, __disp__ - Displacement (cu.in.), __wt__ - Weight (lb/1000), __am__ - Transmission (0 = automatic, 1 = manual).

A exploratory graph (shown in the Appendix) of the influence of three variables, __wt__, __disp__, and __am__ on __mpg__ show depenencies on each.

##A simple linear model is insufficient   


```{r, echo=FALSE, fig.width=5, fig.height=3, fig.align='center'}
library(plyr)
library(ggplot2)
data(mtcars)

p1<-ggplot(mtcars, aes(wt, mpg, color=factor(am))) + geom_point(size=3)
p1<-p1+ggtitle("mpg versus weight")
p1<-p1+labs(x="weight (1000's of lbs)", y="miles per gallon")
p1<-p1+geom_smooth(method=lm)


f1<-lm(mpg~ wt, data=mtcars)
int<-coef(f1)[1]
slp<-coef(f1)[2]


plot_linear_wt<-p1+geom_abline(intercept=int, slope=slp, col="darkmagenta")



```




A model of mpg fitted against weight accounts for about `r round(100*summary(f1)$r.squared, 1)`% of the overall variation. However, there are are systematic variations unaccounted for in the model. This can be seen by the below plot of the residuals, factored for automatic versus manual transmission. 




```{r, echo=FALSE,fig.width=5, fig.height=3}



rmpg<-residuals(f1)
resid<-as.data.frame(rmpg)

mtcars2<-cbind(mtcars2, resid$rmpg)

p<-ggplot(mtcars2, aes(wt, resid$rmpg, color=factor(am))) + geom_point(size=2)
p<-p+ggtitle("Simple Model Residuals")
p<-p+labs(x="weight (1000's of lbs)", y="residuals")
p<-p+geom_smooth(method=lm,se=TRUE)
p<-p+coord_cartesian(ylim = c(-8, 8)) 
p

```
 
The residuals for manual transmission, with > 95% confidence, show a systematic dependency not accounted for in the model.  

##An improved model: including interaction terms and matching the model to physics.

After comparing several models, the model __mpg__ ~ __am__ + __1/ wt__ + __1/disp__ improves the analysis with greatest parsimony. The graph below shows that residuals for the model do not show strong systematic variation with __am__.  

The inverse relationship on __wt__ and __disp__ is what one would expect from the "physics" of the problem since both mass and engine displacment would be expected to relate proportionally to energy and fule consumption.   


```{r, echo=FALSE, fig.align='center', fig.width=5, fig.height=3}
library(plyr)
library(ggplot2)

##plot of data versus inverse weight
p1<-ggplot(mtcars2, aes(invwt, mpg, color=factor(am))) + geom_point(size=2)
p1<-p1+ggtitle("mpg versus 1/weight")
p1<-p1+labs(x="1/weight (1/1000's of lbs)", y="miles per gallon")
p1<-p1+geom_smooth(method=lm)


f2<-lm(mpg~ invwt, data=mtcars2)
int<-coef(f2)[1]
slp<-coef(f2)[2]


plot_inv_wt<-p1+geom_abline(intercept=int, slope=slp, col="darkmagenta")


```


```{r, echo=FALSE, fig.align='center', fig.width=5, fig.height=3}

f3<-lm(mpg~ invwt + invdisp, data=mtcars2)

rmpg3<-residuals(f3)
resid<-as.data.frame(rmpg3)

mtcars2<-cbind(mtcars2, resid$rmpg3)

p<-ggplot(mtcars2, aes(invwt, resid$rmpg3, color=factor(am))) + geom_point(size=2)
p<-p+ggtitle("Model 4 Residuals")
p<-p+labs(x="inverse weight (1000's of lbs)", y="residuals")
p<-p+geom_smooth(method=lm,se=TRUE)
p<-p+coord_cartesian(ylim = c(-8, 8))
p

```

Using "Model 4" below was found to produce the best results with least complexity.

```{r}

f4<-lm(mpg~ factor(am) + invwt+invdisp, data=mtcars2)


```

This model accounts for `r round(100*summary(f4)$r.squared, 1)`% of the variation, a signifcant improvement from the simple model above.  

Based on the fit coefficients, a manual transmission has `r round(summary(f4)$coefficients[2,1],2)` lower miles per gallon, however, the standard error `r round(summary(f4)$coefficients[2,2],2)` has a t-statistic of `r round(summary(f4)$coefficients[2,3],2)` results in a p-value of `r round(summary(f4)$coefficients[2,4],2)` which is above our tolerance of 0.05. Hence, we reject the null hypothesis.  
_Per this analysis there is no statistically significant difference in mpg for manual versus automatic transmissions._
  
    
   
#Appendices

###Exploratory analysis
This graph looks at the dependency of mpg on displacement (__disp__), automatic versus manual transmission(__am__), and vehcile weight (__wt__) in 1000's of lbs.

```{r, fig.align='center', echo=FALSE, fig.width=5}

##plot pairs of displacement, am, and weight
pairs(mpg~disp+am+wt,data=mtcars2, 
   main="Exploratory Analysis")
```

###Plot of a linear model of mpg~wt

A fit of mpg versus weight shows a correlation. Fits for different transmission types do not line up well with the fit of the overall data (shown in magenta).  

```{r, echo=FALSE, fig.align='center', fig.height=3}
plot_linear_wt
```

###Plot of mpg~inverse weight

A fit of mpg versus inverse weight. Fits of the data for different transmissions types line up better with the overall fit (shown in magenta), showing how an improved "physical" interpretation of the model improves overall fit.

```{r, echo=FALSE, fig.align='center', fig.height=3}
plot_inv_wt
```

###Comparison of the simple linear versus Model 4

The Anova shows Model 4 significantly reduces the variances.

```{r, echo=FALSE} 
anova(f1, f4)
```

###Model 4

Summary plot for "Model 4" show the residuals follow a normal distribution to good degree, though some points still could be investigated further. 

```{r, echo=FALSE, fig.height=6}  

summary(f4)

par(mfrow=c(2,2))  
plot(f4)
```

